Moving Elements in List CRDTs

### 2020-02-24 paper/list-move.tex
More consistently use "move" instead of "reorder"

\begin{document}
\title{Reordering Elements in List CRDTs}

@@

\begin{abstract}
Conflict-free Replicated Data Types (CRDTs) for lists allow multiple users to concurrently insert and delete elements in a shared list object.
However, existing algorithms behave poorly when users concurrently reorder elements in the list (i.e.\ move list elements to a new position).
We demonstrate the need for such a move operation, and describe an algorithm that extends a list CRDT with an explicit move operation.
Our algorithm can be used in conjunction with any existing list CRDT algorithm.
In addition to moving a single list element, we also discuss the open problem of moving ranges of elements.
\end{abstract}

@@

All of the aforementioned algorithms allow replicas to \emph{insert} or \emph{delete} elements anywhere in the list.
However, none of them have explicit support for \emph{reordering} or \emph{moving} elements from one position to another position in the list.
This is a surprising omission because reordering is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).

In this paper we introduce an algorithm that allows an existing list CRDT to be extended with a reorder operation.
The algorithm is generic in the sense that it can be implemented on top of any of the aforementioned list CRDTs.

@@

\begin{figure*}
    \centering
    \vspace{3em}
    \begin{tikzpicture}[auto,scale=1.0]
        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,2) -- (17.5,2);
        \node at (1, 4) {Replica $A$:};
        \node at (1, 0) {Replica $B$:};
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 4) {
            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe1) {phone joe}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe2) {phone joe}; \\
        };
        \draw [->,movecolor,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
        \draw [->,movecolor,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
        \draw [->,thick] (6, 0) -- (8.5, 0);
        \draw [->,thick] (6, 4) -- (8.5, 4);
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 4) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \draw [->,thick] (11.5, 0) -- (14.5, 0);
        \draw [->,thick] (11.5, 4) -- (14.5, 4);
        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 4);
        \draw [network] (12.5, 4) to [out=270,in=90] (13.1, 0);
        \node [rotate=90,networkcolor,font=\footnotesize] at (13.5, 2) {network communication};
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 4) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
    \end{tikzpicture}
    \caption{If reordering is implemented by deleting and re-inserting, concurrent moves of the same element result in the anomaly shown here: the moved element is duplicated.}
    \label{fig:duplication}
\end{figure*}

@@

We argue that such duplication in the face of concurrency is undesirable, as it is generally not the behaviour expected by users.
For example, say a user has replicas of their to-do list on their laptop and their smartphone.
They reorder items on their laptop to match new priorities, and also reorder some items on their phone while the phone is offline, as illustrated in Figure~\ref{fig:duplication}.
Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.

@@

\section{A Generic Reordering Algorithm}\label{sec:algorithm}

@@

We will briefly discuss one more approach for moving ranges, and explain why it is not sufficient.
In the example of Figures~\ref{fig:range-bad} and~\ref{fig:range-good}, if the application needs to support reordering of bullet points, we might try to simplify the problem by using a two-level hierarchy instead of a flat sequence of characters.
That is, we can create a top-level list object in which each list element contains the entire content of a single bullet point, and use a separate text CRDT object for the content of each bullet point.
In this design, reordering bullet points requires only a single-element move operation.

@@

Thus, representing each bullet point as a single list element for the sake of reordering does not actually obviate the need for moving ranges of characters.
As before, implementing move as delete-and-reinsert has the effect of duplicating text if two users concurrently split or join the same bullet point.
Thus, it seems that support for moving ranges of characters is an important feature for collaborative document editors.

>>>>

\begin{document}
\title{Moving Elements in List CRDTs}

@@

\begin{abstract}
Conflict-free Replicated Data Types (CRDTs) for lists allow multiple users to concurrently insert and delete elements in a shared list object.
However, existing algorithms behave poorly when users concurrently move list elements to a new position (i.e.\ reorder the list elements).
We demonstrate the need for such a move operation, and describe an algorithm that extends a list CRDT with an explicit move operation.
Our algorithm can be used in conjunction with any existing list CRDT algorithm.
In addition to moving a single list element, we also discuss the open problem of moving ranges of elements.
\end{abstract}

@@

All of the aforementioned algorithms allow replicas to \emph{insert} or \emph{delete} elements anywhere in the list.
However, none of them have explicit support for \emph{moving} elements from one position to another position in the list (\emph{reordering}).
This is a surprising omission because moving is a commonly required operation: in many to-do list applications, a user can drag and drop list elements to reorder them, and graphics software allows users to reorder the object list with commands such as ``bring to front'' (which moves an object to the top of the list, so that it occludes other objects) and ``send to back'' (which moves an object to the bottom, so that it is occluded by other objects).

In this paper we introduce an algorithm that allows an existing list CRDT to be extended with a move operation.
The algorithm is generic in the sense that it can be implemented on top of any of the aforementioned list CRDTs.

@@

\begin{figure*}
    \centering
    \vspace{3em}
    \begin{tikzpicture}[auto,scale=1.0]
        \tikzstyle{item}=[rectangle,draw,inner ysep=4pt,text height=9pt,text depth=2pt,minimum width=2.2cm]
        \tikzstyle{network}=[thick,dashed,networkcolor,-{Stealth[length=3mm]}]
        \path [draw,dotted] (0,2) -- (17.5,2);
        \node at (1, 4) {Replica $A$:};
        \node at (1, 0) {Replica $B$:};
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 4) {
            \node [minimum width=0.5cm] {1.}; & \node (top1) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe1) {phone joe}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (4, 0) {
            \node [minimum width=0.5cm] {1.}; & \node (top2) {buy milk}; \\
            \node [minimum width=0.5cm] {2.}; & \node {water plants}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {3.}; & \node [fill=movecolor!10] (joe2) {phone joe}; \\
        };
        \draw [->,movecolor,thick] (joe1.east) to [in=0, out=0] node [right,near end] {move} (top1.north east);
        \draw [->,movecolor,thick] (joe2.east) to [in=0, out=0] node [right,near end] {move} (top2.north east);
        \draw [->,thick] (6, 0) -- (8.5, 0);
        \draw [->,thick] (6, 4) -- (8.5, 4);
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 4) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (10, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {2.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {3.}; & \node {water plants}; \\
        };
        \draw [->,thick] (11.5, 0) -- (14.5, 0);
        \draw [->,thick] (11.5, 4) -- (14.5, 4);
        \draw [network] (12.5, 0) to [out=90,in=270] (13.1, 4);
        \draw [network] (12.5, 4) to [out=270,in=90] (13.1, 0);
        \node [rotate=90,networkcolor,font=\footnotesize] at (13.5, 2) {network communication};
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 4) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
        \matrix [row sep={19pt,between origins},nodes=item] at (16, 0) {
            \node [minimum width=0.5cm,fill=movecolor!10] {1.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm,fill=movecolor!10] {2.}; & \node [fill=movecolor!10] {phone joe}; \\
            \node [minimum width=0.5cm] {3.}; & \node {buy milk}; \\
            \node [minimum width=0.5cm] {4.}; & \node {water plants}; \\
        };
    \end{tikzpicture}
    \caption{If moving is implemented by deleting and re-inserting, concurrent moves of the same element result in the anomaly shown here: the moved element is duplicated.}
    \label{fig:duplication}
\end{figure*}

@@

We argue that such duplication in the face of concurrency is undesirable, as it is generally not the behaviour expected by users.
For example, say a user has replicas of their to-do list on their laptop and their smartphone.
They move items on their laptop to match new priorities, and also move some items on their phone while the phone is offline, as illustrated in Figure~\ref{fig:duplication}.
Later, when the phone is online again and synchronises with the laptop, any items that were moved on both devices will be duplicated.

@@

\section{A Generic Moving Algorithm}\label{sec:algorithm}

@@

We will briefly discuss one more approach for moving ranges, and explain why it is not sufficient.
In the example of Figures~\ref{fig:range-bad} and~\ref{fig:range-good}, if the application needs to support reordering of bullet points, we might try to simplify the problem by using a two-level hierarchy instead of a flat sequence of characters.
That is, we can create a top-level list object in which each list element contains the entire content of a single bullet point, and use a separate text CRDT object for the content of each bullet point.
In this design, moving bullet points requires only a single-element move operation.

@@

Thus, representing each bullet point as a single list element for the sake of moving does not actually obviate the need for moving ranges of characters.
As before, implementing move as delete-and-reinsert has the effect of duplicating text if two users concurrently split or join the same bullet point.
Thus, it seems that support for moving ranges of characters is an important feature for collaborative document editors.
